\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{hyperref}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage[preprint]{icml2026}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}

\icmltitlerunning{PubMedQA v1 Local Evidence Baseline}

\begin{document}

\twocolumn[
\icmltitle{PubMedQA v1: Local Evidence-Aware Biomedical Question Answering over PubMed}

\begin{icmlauthorlist}
\icmlauthor{Timothy Daley}{pubmedqa}
\end{icmlauthorlist}

\icmlaffiliation{pubmedqa}{Personal research project}
\icmlcorrespondingauthor{Timothy Daley}{timothydaley@example.com}

\icmlkeywords{biomedical retrieval, PubMed, local language models, retrieval-augmented generation, evaluation}

\vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
This report updates the PubMedQA local baseline after the v1 retrieval, answer-generation, and release-gate work. The system combines MedCPT dense retrieval, SQLite FTS5 lexical retrieval, reciprocal-rank fusion, bounded evidence/citation/recency scoring, duplicate suppression, clinical-query-aware context balancing, local MLX generation, second-pass evidence summaries, and PMID citation validation. On a 13-question expanded benchmark, the current citation+recency retrieval baseline returns 130 papers with zero Qwen-judged irrelevant papers, 129/130 papers judged relevance $\geq 2$, and mean judged relevance 2.577/3. The regenerated answer benchmark includes an explicit Evidence basis section in all 13 answers and has zero invalid PMID citations or citation-extraction mismatches. We define v1 release gates covering build integrity, retrieval quality, answer quality, latency, and manual edge-case review.
\end{abstract}

\section{Purpose}

The project goal is to answer biomedical questions using local PubMed evidence rather than web search or remote retrieval. A local snapshot stores PubMed metadata in SQLite with FTS5 and MedCPT article vectors in LanceDB. Query-time execution is designed to be offline except for optional enrichment and evaluation-time PubMed fetching. This document records the current v1-oriented baseline and supersedes the earlier five-question v4 report.

\section{System}

Given a question, the system embeds the query with the MedCPT query encoder \citep{jin2023medcpt}, retrieves candidates from a LanceDB vector table \citep{lancedb}, retrieves lexical candidates from SQLite FTS5, and fuses dense and lexical ranks with reciprocal rank fusion \citep{rrf}. Retrieved records are then rescored by a bounded additive score:
\begin{equation}
  s_{final} = s_{rel} + \alpha I + B_{overlap} + B_{focus} + B_{evidence},
\end{equation}
where $s_{rel}$ is normalized fused relevance, $I$ is a bounded impact term derived from RCR, log citation count, citation rate, and recency when available \citep{icite}, and the remaining terms reward query overlap, LLM-extracted focus-keyword overlap, and evidence type. The final context is deduplicated by title similarity and balanced between clinical/review and mechanistic/other papers.

Generation uses a local MLX language model \citep{mlx}. Answers are prompted to cite every factual claim with retrieved PMIDs, avoid causal language for observational/mechanistic evidence, distinguish endpoint classes, and include a concise \texttt{Evidence basis} section. A citation validator scans the generated answer and reports any PMID not present in the retrieved context.

The CLI now also supports optional second-pass evidence synthesis. Users can summarize either all retrieved papers or only cited papers, use direct or map-reduce summarization, and export per-paper evidence notes for debugging and factual review.

\section{Recent Changes}

The current baseline includes five major changes since the earlier v4 paper.

\paragraph{Citation and recency scoring.} Retrieval impact scoring now combines RCR/log citation count, citation velocity, and recency. This keeps highly cited established papers competitive while allowing recent papers to surface when citation velocity or recency is informative.

\paragraph{Clinical context balancing.} Clinical/outcome-oriented questions prefer clinical/review evidence when enough candidates exist. The usual clinical query target is seven clinical/review papers and three mechanistic/other papers; mechanistic questions can still return fully mechanistic contexts.

\paragraph{Answer prompt calibration.} The answer prompt now explicitly asks for an \texttt{Evidence basis} section and cautions against unqualified ``yes'' answers for observational associations. It instructs the model to distinguish incidence/prevention, survival after diagnosis, treatment response, surrogate outcomes, and mechanistic findings.

\paragraph{Answer comparison tooling.} A new \texttt{scripts/compare\_answers.py} script compares answer artifacts, including latency, citation validity, citation-set changes, Evidence basis coverage, citation-status mismatches, and simple language review flags.

\paragraph{Build integrity and release gates.} Snapshot builds now report duplicate PMIDs, LanceDB row counts, SQLite row counts, abstract/year/citation coverage, and integrity checks. The project now has explicit v1 release gates for build integrity, retrieval quality, answer quality, latency, and manual review.

\section{Benchmark}

We use the expanded 13-question benchmark in \texttt{benchmarks/expanded\_questions.txt}. It covers drug safety/effectiveness, prevention, supplementation, cardiovascular and kidney outcomes, lifestyle interventions, mechanistic biology, and prognosis.

The current key artifacts are the citation+recency retrieval baseline, the strict Qwen judged retrieval baseline, the Evidence-basis answer baseline, the citation/recency review note, and the v1 release-gate checklist. Full artifact filenames are recorded in the repository and release notes.

The retrieval judge is \texttt{mlx-community/Qwen3-4B-4bit}, run locally through MLX with a strict JSON rubric. The answer generator is the configured local MLX model.

\section{Retrieval Results}

\begin{table*}[t]
\centering
\caption{Current citation+recency retrieval baseline on the expanded benchmark. Direct@10 is the number of papers Qwen judged relevance 3/3. Rel@10 is the number judged at least 2/3.}
\label{tab:retrieval}
\begin{tabular}{lrrrrr}
\toprule
Question & Retrieval (s) & Clinical & Mech. & Direct@10 & Rel@10 \\
\midrule
Metformin / cancer risk & 14.56 & 7 & 3 & 2 & 10 \\
Statins / dementia & 6.37 & 7 & 3 & 8 & 10 \\
Vitamin D / respiratory infections & 12.80 & 7 & 3 & 4 & 10 \\
Aspirin / primary CVD prevention & 20.54 & 7 & 3 & 7 & 10 \\
GLP-1 / cardiovascular outcomes & 13.79 & 7 & 3 & 10 & 10 \\
Beta blockers / heart failure mortality & 13.75 & 7 & 3 & 9 & 10 \\
SGLT2 inhibitors / kidney outcomes & 13.24 & 7 & 3 & 8 & 9 \\
Omega-3 / cardiovascular prevention & 13.29 & 7 & 3 & 6 & 10 \\
Exercise / insulin sensitivity & 13.64 & 5 & 5 & 3 & 10 \\
Ketogenic diet / adult epilepsy & 11.56 & 5 & 5 & 3 & 10 \\
Aspirin / colorectal cancer risk & 10.61 & 7 & 3 & 6 & 10 \\
Metformin / AMPK mechanism & 14.80 & 0 & 10 & 2 & 10 \\
p53 mutation / cancer prognosis & 12.98 & 2 & 8 & 8 & 10 \\
\midrule
Mean / total & 13.22 & 5.8 & 4.2 & 76/130 & 129/130 \\
\bottomrule
\end{tabular}
\end{table*}

Table~\ref{tab:retrieval} shows that the current retrieval baseline has no judged irrelevant papers in the top ten across the benchmark. Average mean judge relevance is 2.577/3. Compared with the previous entity-fix baseline, citation+recency scoring changed top-10 membership for only 4/13 questions, preserved an average PMID-set Jaccard overlap of 0.909, increased total Direct@10 papers from 75 to 76, and kept Rel@10 at 129/130. The principal cost is latency: average retrieval time increased from 11.26s to 13.22s in the judged runs.

\section{Answer Generation Results}

\begin{table*}[t]
\centering
\caption{Current answer baseline with the Evidence basis prompt. Citation validation checks whether generated PMIDs are in the retrieved context.}
\label{tab:answers}
\begin{tabular}{lrrrr}
\toprule
Question & Retrieval (s) & Generation (s) & Cited PMIDs & Invalid citations \\
\midrule
Metformin / cancer risk & 15.00 & 10.05 & 4 & 0 \\
Statins / dementia & 6.89 & 8.89 & 3 & 0 \\
Vitamin D / respiratory infections & 14.84 & 8.80 & 5 & 0 \\
Aspirin / primary CVD prevention & 23.70 & 6.61 & 5 & 0 \\
GLP-1 / cardiovascular outcomes & 16.17 & 9.84 & 4 & 0 \\
Beta blockers / heart failure mortality & 15.89 & 8.22 & 3 & 0 \\
SGLT2 inhibitors / kidney outcomes & 14.73 & 9.65 & 4 & 0 \\
Omega-3 / cardiovascular prevention & 13.81 & 9.22 & 3 & 0 \\
Exercise / insulin sensitivity & 12.83 & 10.77 & 5 & 0 \\
Ketogenic diet / adult epilepsy & 10.90 & 7.99 & 4 & 0 \\
Aspirin / colorectal cancer risk & 12.25 & 11.29 & 5 & 0 \\
Metformin / AMPK mechanism & 15.36 & 10.71 & 4 & 0 \\
p53 mutation / cancer prognosis & 14.08 & 11.14 & 4 & 0 \\
\midrule
Mean & 14.34 & 9.47 & 4.1 & 0 \\
\bottomrule
\end{tabular}
\end{table*}

The regenerated answer benchmark fixed the main prompt-structure gap. Relative to the previous prompt-fix artifact, Evidence basis coverage improved from 0/13 to 13/13. Citation validation remained clean: zero citation warnings, zero invalid cited-PMID rows, and zero manual citation extraction mismatches. Simple review flags in \texttt{compare\_answers.py} dropped from 13 rows to 0 rows because the missing Evidence basis issue was resolved.

\section{Known Limitations}

First, the local judge is useful for repeatable regression checks but is not a substitute for expert biomedical review. It can be generous for reviews, editorials, and indirect evidence. Second, some clinical questions still retrieve endpoint-mismatched papers, such as survival-after-diagnosis papers for prevention/incidence questions. Third, population constraints require more tuning: for example, adult epilepsy queries can still surface pediatric ketogenic diet papers. Fourth, citation and recency features are useful but not sufficient; stale mechanistic papers can still enter clinical contexts when lexical/vector similarity is high. Fifth, the system retrieves abstracts and metadata, not full text.

\section{Reproduction Commands}

The current artifacts were produced with the following command pattern, using the expanded benchmark question file and the artifact names listed in the repository:
\begin{footnotesize}
\begin{verbatim}
export PUBMEDQA_DATA="/path/to/pubmed_qa"

python eval.py --questions expanded.txt \
  --out RETRIEVAL_JSON --print-titles

python judge.py --input RETRIEVAL_JSON \
  --out JUDGED_JSON --provider mlx \
  --model QWEN3_4B_4BIT

python eval.py --questions expanded.txt \
  --with-llm --out ANSWERS_JSON

python compare_answers.py OLD NEW
\end{verbatim}
\end{footnotesize}

\section{v1 Release Gates}

The v1 release gates require: build-integrity checks to pass or have documented non-blocking warnings; 13/13 benchmark questions returning ten results; no judged irrelevant top-10 papers or a documented exception; at least 125/130 papers judged relevance $\geq 2$; average mean judge relevance at least 2.50; 13/13 answers with an Evidence basis section; zero invalid citations; average retrieval latency below 20s; average answer generation latency below 15s; and manual review of known edge cases including metformin/cancer, aspirin/CRC, aspirin primary prevention, exercise/insulin sensitivity, and p53/prognosis.

The current baseline passes the measured retrieval, answer, and latency gates. Remaining work before tagging v1 is primarily manual review plus targeted fixes for endpoint, population, and stale-mechanistic retrieval issues.

\section{Next Steps}

Near-term work should be targeted rather than broad enrichment: verify the improved build-integrity report on a fresh small snapshot; add a stale-mechanistic penalty for clinical outcome questions; improve population/age filtering; strengthen endpoint matching for prevention versus post-diagnosis survival; and use the v1 release gates as the decision checklist for freezing the first release.

\bibliography{pubmedqa_v4_baseline_icml2026}
\bibliographystyle{icml2026}

\end{document}
